### 2022-10-14 chapters/introduction.tex
add : notes to be discussed in the future works
update: introduction

While there are many fields where one can optimize energy consumption.
Our focus is on the \emph{Information and Communication Technology} (ICT) sector, which is expected to account for around 4\% of global \emph{GreenHouse Gas} (GHG) emissions in 2020, with an alarming 8\% growth rate, according to the French think tank The Shift Project.\footnote{\url{ https://www.theshiftproject.org/article/ict-environmental-impact/}}
According to \emph{Statista}, the energy consumption of ICT increased from 4.3 exajoules in 2018 to 5.8 exajoules in 2025.\footnote{\url{https://www.statista.com/statistics/271139/energy-consumption-of-ict-worldwide}}

\begin{figure}[!h]
    \centering
    \includegraphics[width=.7\linewidth]{chapters/distribution_of_ict_consumption.png}
    \caption{Final energy consumption of digital technologies by item in 2017}
    \label{fig:distribution_of_ict_consumption}
\end{figure}

Figure~\ref{fig:distribution_of_ict_consumption} shows the distribution of ICT consumption in 2018, where the largest chunk of the energy consumption is due to data usage, aka 55\% of the energy consumed where 35\% of this energy is used by data centers.
Reducing energy consumption means reducing the impact 19\% of the ICT energy consumption has on the environment.

@@

The best way to do so is to formulate a theory behind the energy consumption of algorithms, such as the complexity and the o notation.
Unfortunately, this is not possible in the current state of the art.
Due to the lack of knowledge about the energy consumption of the algorithms, and the strong correlation between this consumption and the hardware configuration.
Unlike algorithm optimization in the field of performance, which is agnostic toward the platform, the energy consumption of the algorithms is dependent on the execution environment.
Therefore, for the moment, we will start by formulating some hypotheses and exploring them using empirical analysis.
Figure~\ref{fig:thesis_position} highlights this thesis's position on the sustainability of ICT\@; while this thesis only addresses a small portion of ICT's energy usage, we feel it is a step in the right direction for additional solutions to mature to preserve humanity.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.7\textwidth,height=\textheight,keepaspectratio]{chapters/thesis_position.pdf}
    \caption{Our position in the IT sustainability research}
    \label{fig:thesis_position}
\end{figure}
\subsection{Objectives}
The purpose of this thesis is to help developers build more energy-efficient software.
Unfortunately, when it comes to the energy consumption of programs, there is a lack of awareness and knowledge among software developers~\cite{ournani2020reducing,pang2015programmers,pinto2014mining}.
This is mainly due to the lack of tools that can help developers understand the energy consumption of their programs.
Therefore, we aim to provide clear, understandable, and easy-to-use tools and guidelines that can help developers reduce the energy consumption of their programs.

@@

The final strategy consists of \emph{deliberately reducing the performance of the software to reduce its energy consumption}.
The goal of this method is to reduce energy usage by sacrificing execution time.
This strategy is useful while dealing with services since the task has an unlimited execution time.
The second part is to use some. opportunistic schedulers, which will stop the software if it is estimated to spend more energy.
This strategy will be discussed first in Chapter~\ref{chapter:porgramming_langauges}, then we will continue with this strategy in the perspectives section.

\subsection{Organization}
\begin{figure}[!h]
    \centering
    \includegraphics[width=.7\textwidth,height=\textheight,keepaspectratio]{chapters/thesis_contributions.pdf}
    \caption{Contributions reported in this thesis}
    \label{fig:thesis_contributions}
\end{figure}

Figure~\ref{fig:thesis_contributions} summarizes the work completed for this thesis As mentioned in the previous section, our goal is to help developers reduce the energy consumption of their programs during execution.
We will use an empirical approach to accomplish this.
As a result, we will require some tools to assist us in this task.
To begin, we will provide a tool to assist developers in measuring the energy consumption of their programs.
The rest of the manuscript is outlined below.

@@

    \item \cref{chapter:literature_review} provides a literature review of the energy consumption of software.

@@

          In this section, we provide the literature approaches used to measure energy consumption and classify them, compare them and discuss their pros and cons when used for our research. After that, we will tackle the challenge of accuracy within the field of energy consumption, and discuss how the literature tried to overcome these challenges. Finally, we will present some of the most recent works in the field of software energy consumption optimization;

@@

    \item \cref{chapter:benchmarking} will discuss two facets of empirical tests and how we adapt them to energy consumption measurements.
          First, we will discuss the reproducibility challenge and how to overcome it with the usage of containers, then we will push this approach further by proposing and new protocol to make tests not only reproducible but able to be extended with other features to overcome the rapid pace of software evolution.
          The second part of this chapter will discuss the aspect of accuracy when it comes to energy consumption measurements.

@@

          Then, it will provide some solutions to reduce this energy variation and improve the accuracy of the measurements;
    \item After setting the ground for energy benchmarks in the previous chapter, we discuss energy consumption one of the most popular programming languages in the world, Python, in \cref{chapter:python}.
          This chapter reports on how much python costs in terms of energy consumption compared to other languages.

@@

          We will do this by using a different implementation of the Python interpreter.

@@

          In \cref{chapter:java}, we try to optimize the java code, using by changing the default JVM implementation.
          We will compare the energy consumption of $12$ benchmarks using $52$ implementation, each benchmark is dedicated to a typical use case.

@@

    \item In contrast, \cref{chapter:porgramming_langauges} will use the flexibility provided by the micro-services architecture~\cite{dmitry2014micro} to analyze each programming language's energy behavior in light of various web scenarios to optimize the energy use of web services.
          We will first examine the effects of the various programming languages when dealing with the \emph{Remote Procedure Call} (RPC) Protocol.
          In this instance, we will use two scaling factors: the number of concurrent clients and the size of the requests. Then, we will compare $261$ web frameworks, each implementing the same website using seven use cases.

@@

    \item Finally, we conclude our work in \cref{chapter:conclusion} by summarizing the work done in this thesis and discussing the future work that can be done to further reduce the energy consumption of software.

@@

\subsection*{Tools}
\begin{itemize}
    \item \bibentry{Belgaid_JouleHunter_an_2021},
    \item \bibentry{Belgaid_JRefferal_Which_JVM_2021},
    \item \bibentry{Belgaid_Jouleit_a_2020},
          % \item \bibentry{Belgaid_Ijoules_Python_library_2020}
    \item \bibentry{Belgaid_Pyjoules_Python_library_2019}.
\end{itemize}

% \subsection*{Posters}
% \begin{itemize}
%     \item \emph{PyRAPL\@: Make your python green again}
%           % TODO : ask romain about the date and the name of the event 
% \end{itemize}
\subsection*{Future Work}
\begin{itemize}
    \item Reducing the energy consumption of Python using non-intrusive techniques,
    \item Empirical analysis on the energy consumption of different web frameworks,
    \item The impact of programming languages on energy consumption of web services (a case study of RPC protocol),
    \item THow do ORMs affect how much energy is used? A case study of Django and Flask.
\end{itemize}

\cleardoublepage

>>>>

While there are many fields where one can optimize energy consumption.
Our focus is on the \emph{Information and Communication Technology} (ICT) sector, which is expected to account for around 4\% of global \emph{GreenHouse Gas} (GHG) emissions in 2020, with an alarming 8\% growth rate, according to the French think tank The Shift Project.\footnote{\url{ https://www.theshiftproject.org/article/ict-environmental-impact/}}
According to \emph{Statista}, the energy consumption of ICT increased from 4.3 exajoules in 2018 to 5.8 exajoules in 2025.\footnote{\url{https://www.statista.com/statistics/271139/energy-consumption-of-ict-worldwide}}
\begin{figure}[!h]
    \centering
    \includegraphics[width=.7\linewidth]{chapters/distribution_of_ict_consumption.png}
    \caption{Final energy consumption of digital technologies by item in 2019(The Shift Project – Forecast Model 2021\cite{shift_2021})}
    \label{fig:distribution_of_ict_consumption}
\end{figure}

Figure~\ref{fig:distribution_of_ict_consumption} shows the distribution of ICT consumption in 2018, where the largest chunk of the energy consumption is due to data usage, aka 63\% of the energy consumed where 22\% of this energy is used by data centers.
Reducing energy consumption means reducing the impact 14\% of the ICT energy consumption has on the environment.

@@

The best way to do so is to formulate a theory behind the energy consumption of algorithms, such as the complexity and the o notation.
Unfortunately, this is not possible in the current state of the art.
Due to the lack of knowledge about the energy consumption of the algorithms, and the strong correlation between this consumption and the hardware configuration.
Unlike algorithm optimization in the field of performance, which is agnostic toward the platform, the energy consumption of the algorithms is dependent on the execution environment.
Therefore, for the moment, we start by formulating some hypotheses and exploring them using empirical analysis.
Figure~\ref{fig:thesis_position} highlights this thesis's position on the sustainability of ICT\@; while this thesis only addresses a small portion of ICT's energy usage, we feel it is a step in the right direction for additional solutions to mature to preserve humanity.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.7\textwidth,height=\textheight,keepaspectratio]{chapters/thesis_position.pdf}
    \caption{Our position in the IT sustainability research}
    \label{fig:thesis_position}
\end{figure}
\section{Objectives}
The purpose of this thesis is to help developers build more energy-efficient software.
Unfortunately, when it comes to the energy consumption of programs, there is a lack of awareness and knowledge among software developers~\cite{ournani2020reducing,pang2015programmers,pinto2014mining}.
This is mainly due to the lack of tools that can help developers understand the energy consumption of their programs.
Therefore, we aim to provide clear, understandable, and easy-to-use tools and guidelines that can help developers reduce the energy consumption of their programs.

@@

The final strategy consists of \emph{deliberately reducing the performance of the software to reduce its energy consumption}.
The goal of this method is to reduce energy usage by sacrificing execution time.
This strategy is useful while dealing with services since the task has an unlimited execution time.
The second part is to use some. opportunistic schedulers, which will stop the software if it is estimated to spend more energy.
This strategy is discussed first in Chapter~\ref{chapter:porgramming_langauges}, then we continue with this strategy in the perspectives section.

\section{Organization}
\begin{figure}[!h]
    \centering
    \includegraphics[width=.7\textwidth,height=\textheight,keepaspectratio]{chapters/thesis_contributions.pdf}
    \caption{Contributions reported in this thesis}
    \label{fig:thesis_contributions}
\end{figure}

Figure~\ref{fig:thesis_contributions} summarizes the work completed for this thesis As mentioned in the previous section, our goal is to help developers reduce the energy consumption of their programs during execution.
We use an empirical approach to accomplish this.
As a result, we require some tools to assist us in this task.
To begin, we provide a tool to assist developers in measuring the energy consumption of their programs.
The rest of the manuscript is outlined below.

@@

    \item Chapter~\ref{chapter:literature_review} provides a literature review of the energy consumption of software.

@@

          In this section, we provide the literature approaches used to measure energy consumption and classify them, compare them and discuss their pros and cons when used for our research. After that, we tackle the challenge of accuracy within the field of energy consumption, and discuss how the literature tried to overcome these challenges. Finally, we  present some of the most recent works in the field of software energy consumption optimization;

@@

    \item Chapter~\ref{chapter:benchmarking} discusses two facets of empirical tests and how we adapt them to energy consumption measurements.
          First, we discuss the reproducibility challenge and how to overcome it with the usage of containers, then we push this approach further by proposing and new protocol to make tests not only reproducible but able to be extended with other features to overcome the rapid pace of software evolution.
          The second part of this chapter discusses the aspect of accuracy when it comes to energy consumption measurements.

@@

          Then, it provides some solutions to reduce this energy variation and improve the accuracy of the measurements;
    \item After setting the ground for energy benchmarks in the previous chapter, we discuss the energy consumption of Python, one of the world's most popular programming languages.
          Chapter~\ref{chapter:python} reports on how much python costs in terms of energy consumption compared to other languages.

@@

          We achieve this by using a different implementation of the Python interpreter.

@@

          In Chapter~\ref{chapter:java}, we try to optimize the Java code, using by changing the default JVM implementation.
          We compare the energy consumption of $12$ benchmarks using $52$ implementation, each benchmark is dedicated to a typical use case.

@@

    \item In contrast, Chapter~\ref{chapter:porgramming_langauges}  uses the flexibility provided by the micro-services architecture~\cite{dmitry2014micro} to analyze each programming language's energy behavior in light of various web scenarios to optimize the energy use of web services.
          We first examine the effects of the various programming languages when dealing with the \emph{Remote Procedure Call} (RPC) Protocol.
          In this instance, we use two scaling factors: the number of concurrent clients and the size of the requests. Then, we compare $261$ web frameworks, each implementing the same website using seven use cases.

@@

    \item Finally, we conclude our work in Chapter~\ref{chapter:conclusion} by summarizing the work done in this thesis and discussing the future work that can be done to further reduce the energy consumption of software.

@@

\subsection*{Tools}
% \begin{itemize}
%     \item \textbf{Jouleit} (\url{github.com/powerapi-ng/jouleit}):  \bibentry{Belgaid_JouleHunter_an_2021},
%     \item \bibentry{Belgaid_JRefferal_Which_JVM_2021},
%     \item \bibentry{Belgaid_Jouleit_a_2020},
%           % \item \bibentry{Belgaid_Ijoules_Python_library_2020}
%     \item \bibentry{Belgaid_Pyjoules_Python_library_2019}.
% \end{itemize}
\begin{itemize}
    \item \textbf{Jouleit} (\url{github.com/powerapi-ng/jouleit}): a tool that can be used to monitor energy consumption for any Linux program, this tool was used to compare the energy consumption of different JVMs;
          \\
          \bibentry{Belgaid_Jouleit_a_2020},
    \item \textbf{JRefferal} (\url{github.com/chakib-belgaid/jreferral}): a tool that allows the user to explore the JVM settings and their impact on the energy consumption of a given Java program. This tool was the result of the second article of this thesis;
          \\
          \bibentry{Belgaid_JRefferal_Which_JVM_2021},
    \item \textbf{PyJoules} (\url{pypi.org/project/pyJoules}) is a software toolkit to measure the energy footprint of a host machine along the execution of a piece of Python code. It can measure the energy consumption on the level of script, function, and bloc of code;
          \\
          \bibentry{Belgaid_Pyjoules_Python_library_2019},
    \item \textbf{JouleHunter} (\url{pypi.org/project/joulehunter}): an energy profiler for python applications. It can be used to highlight the functions that consume the most energy in a given Python program. Its main usage is to help developers do an exploratory analysis of their application to scope the functions that should be optimized to be then targeted by PyJoules;
          \\
          \bibentry{Belgaid_JouleHunter_an_2021},
    \item \textbf{GreenBoard} (\url{github.com/chakib-belgaid/greenboard}): a dashboard designed to help developers choose the best stack for their web application. It is based on the results of the third article of the last chapter.
\end{itemize}

% \subsection*{Future Work}
% \begin{itemize}
%     \item Reducing the energy consumption of Python using non-intrusive techniques,
%     \item Empirical analysis on the energy consumption of different web frameworks,
%     \item The impact of programming languages on energy consumption of web services (a case study of RPC protocol),
%     \item How do ORMs affect how much energy is used? A case study of Django and Flask.
% \end{itemize}

% \cleardoublepage

### 2022-10-14 chapters/literature/draft_citations.tex
update : centring graphics
add : introduction
update: introduction

\newpage
\section{Benchmarking}\label{sec:soa_benchmarking}
This section will go through the flaws and best practices of empirical research before applying them to our topic of study.

@@

\begin{figure}[!hbt]
    \centering{\includegraphics{imgs/power-planes-rapl}}
    \caption{INTEL RAPL scopes
    }\label{fig:rapl-domaine}
\end{figure}

@@

The first examples of software measurement tools are PowerAPI~\cite{colmant2018next}, SmartWatts formula~\cite{fieni2020smartwatts} and SelfWatts~\cite{fieni2021selfwatts}.
These tools collect global energy consumption measurements from RAPL and use other system events such as cache misses/hits and CPU frequency evolution (DVFS) via a sensor to construct a power model of the control groups (system control groups, docker containers, Kubernetes pods, etc.) using a Ridge regression.
The model continuously learns and improves its real-time energy usage data precision with a maximum frequency of 100 Hz.
The instrument has a decentralized, lightweight design.
Only the lightest sensors required for data collection and transmission are put into the monitored devices.
The SmartWatts formula is then executed on the primary server to construct the model that permits assigning the energy usage for each functional control group.
PowerAPI is only compatible with Linux on a bare-metal physical computer.
\\
WattWatcher~\cite{lebeane2015watt} is a multi-core power measuring framework that provides process-level energy measurements.
This program uses power models to predict process energy usage. It uses CPU events passed from the measured node to a model generator node to construct the power model. It works by combining a description of the CPU with a list of the hardware events through multiple calibration phases to build a robust model.

@@

\section{Conclusion}
As we have seen in the state of the art, many methods have been proposed to reduce the ICT's energy footprint, which can be applied in different parts of the program's lifecycle, from consumption to execution.
Furthermore, the execution phase took the attention of many researchers because it is the part where the most energy is consumed.
This thesis will focus on that aspect as well.
However, unlike most work done on the hardware aspect, we will target the software impact on this energy, starting from the choice of the programming language to how to tune some features of a framework to make the software consumes less energy.
To do so, we use the empirical approach due to its consistency for the moment.
Unlike the performance, which is essentially related to the algorithm's complexity, the energy consumption is more impacted by the hardware.
Therefore, to optimize the energy consumption, we choose a spiral method based on 3 phases:
\begin{enumerate}
    \item executes the code,
    \item measure the program,
    \item infers the guidelines.
\end{enumerate}

>>>>

\clearpage
\section{Benchmarking}\label{sec:soa_benchmarking}
This section will go through the flaws and best practices of empirical research before applying them to our topic of study.

@@

\begin{figure}[!hbt]
    \centering
    \includegraphics[width=0.8\linewidth]{imgs/power-planes-rapl}
    \caption{INTEL RAPL scopes
    }\label{fig:rapl-domaine}
\end{figure}

@@

The first examples of software measurement tools are PowerAPI~\cite{colmant2018next}, SmartWatts formula~\cite{fieni2020smartwatts} and SelfWatts~\cite{fieni2021selfwatts}.
These tools collect global energy consumption measurements from RAPL and use other system events such as cache misses/hits and CPU frequency scaling (DVFS) via a sensor to construct a power model of the control groups (system control groups, docker containers, Kubernetes pods, etc.) using a Ridge regression.
The model continuously learns and improves its real-time energy usage data precision with a maximum frequency of 100 Hz.
The instrument has a decentralized, lightweight design.
Only the lightest sensors required for data collection and transmission are put into the monitored devices.
The SmartWatts formula is then executed on the primary server to construct the model that permits assigning the energy usage for each functional control group.
PowerAPI is only compatible with Linux on a bare-metal physical computer.
\\
WattWatcher~\cite{lebeane2015watt} is a multi-core power measuring framework that provides process-level energy measurements.
This program uses power models to predict process energy usage. It uses CPU events passed from the measured node to a model generator node to construct the power model. It works by combining a description of the CPU with a list of the hardware events through multiple calibration phases to build a robust model.

@@

\section{Conclusion}
As we have seen in the state of the art, many methods have been proposed to reduce the ICT's energy footprint, which can be applied in different parts of the program's lifecycle, from consumption to execution.
Furthermore, the execution phase took the attention of many researchers because it is the part where the most energy is consumed.
This thesis will focus on that aspect as well.
However, unlike most work done on the hardware aspect, we will target the software impact on this energy, starting from the choice of the programming language to how to tune some features of a framework to make the software consumes less energy.
To do so, we use the empirical approach due to its consistency for the moment.
Unlike the performance, which is essentially related to the algorithm's complexity, the energy consumption is more impacted by the hardware.
Therefore, to optimize energy consumption, we choose a spiral method based on 3 phases:
\begin{enumerate}
    \item execute the code,
    \item measure program's energy consumption,
    \item infer the guidelines.
\end{enumerate}
